Approximation as Differentiation

### 2025-07-10 main/conclusion.tex
Tweaks.

\paragraph{Quantitative Galois slicng and XAI.}  Explainable AI (XAI) techniques like Gradient-weighted Class
Activation Mapping (Grad-CAM)~\cite{selvaraju20} use reverse-mode AD selectively to calculate \emph{heat maps}
(or \emph{saliency maps}) that highlight input regions contributing to a given classification or other
outcome. We would like to investigate quantitative approximation structures where $\top$ represents the
original input image and lower elements represent ``slices'' of the image where individual pixels have been
ablated to some degree (partly removed or blurred). This might allow for composing some of these techniques
with Galois slicing, for use in hybrid systems such as physical simulations with ML-based parameterisations.

@@

\paragraph{Alternative Higher-order Models}

@@

\paragraph{Implementation}

@@

An interesting implementation technique, as an alternative the denotational approach presented here or the
trace-based approaches used in earlier \GPS implementations, would be to develop a source-to-source
transformation, in direct analogy with the CHAD approach to automatic differentiation
\cite{vakar22,nunes2023}. In their approach, forward and reverse-mode AD are implemented as compositional
transformations on source code, guided by a universal property: they arise as the unique structure-preserving
functors from the source language to a suitably structured target language formalised as a Grothendieck
construction. Adapting this to Galois slicing would allow slicing to ``compiled in'', avoiding the need for a
custom interpreter and presenting opportunities for optimisation.

>>>>

\paragraph{Quantitative slicing and XAI}  Explainable AI (XAI) techniques like Gradient-weighted Class
Activation Mapping (Grad-CAM)~\cite{selvaraju20} use reverse-mode AD selectively to calculate \emph{heat maps}
(or \emph{saliency maps}) that highlight input regions contributing to a given classification or other
outcome. We would like to investigate quantitative approximation structures where $\top$ represents the
original input image and lower elements represent ``slices'' of the image where individual pixels have been
ablated to some degree (partly removed or blurred). This might allow for composing some of these techniques
with Galois slicing, for use in hybrid systems such as physical simulations with ML-based parameterisations.

@@

\paragraph{Alternative Higher-Order Models}

@@

make\paragraph{General Inductive and Coinductive Types}

\paragraph{Source-To-Source Translation Techniques}

@@

An interesting alternative to the denotational approach presented here, and to the trace-based approaches used
in earlier \GPS implementations, would be to develop a source-to-source transformation, in direct analogy with
the CHAD approach to automatic differentiation \cite{vakar22,nunes2023}. In their approach, forward and
reverse-mode AD are implemented as compositional transformations on source code, guided by a universal
property: they arise as the unique structure-preserving functors from the source language to a suitably
structured target language formalised as a Grothendieck construction. Adapting this to Galois slicing would
allow slicing to ``compiled in'', avoiding the need for a custom interpreter and potentially exposing
opportunities for optimisation.
